Linear and Differential Cryptanalysis of DES

### 2009-01-30 trunk/talk/draft1.tex
DES Linear slides complete

\subsection{Linear Attack: 3-round DES}
\begin{frame}
\begin{itemize}
\item{Structurally and cryptographically, DES is similar to SPN.}
\item{We will adapt the linear attack on SPN to a linear attack on a 3-round DES algorithm and mount this attack.}
\item{We will also show how to extend this attack to a full 16-round DES algorithm and show it is faster than an exhaustive search.}
\end{itemize}
\end{frame}

@@

>>>>

\subsection{Linear Attack: 4-round DES}
\begin{frame}
\begin{itemize}
\item{Structurally and cryptographically, DES is similar to SPN.}
\item{We will adapt the linear attack on SPN to a linear attack on a 3-round DES algorithm and mount this attack.}
\item{We will also show how to extend this attack to a full 16-round DES algorithm and show it is faster than an exhaustive search.}
\end{itemize}
\end{frame}

@@

\begin{frame}
We perform the attack on the 4-round DES similiarly to how we performed the attack on 4-round SPN, by "undoing" the last round.\\
\vspace{5mm}
\pause
Note that in the last round, $C_L$ bits are not changed, only $C_R$ bits are affected. Note also that there is only one $C_R$ bit the 3-round equation. I
\vspace{5mm}
\pause
This bit, $C_R{17}$ is affected by the result of S-Box 1 in round 4. If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
\vspace{5mm}
\pause
Thus, this attack will give us $6$ of the $56$ key bits. What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^56$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^50$ bits.
\end{frame}

\begin{frame}
How many pairs do we need? Probabalistic justification:\\
\vspace{3mm}
Matsui Lemma 5: Let $N$ be the number of given random plaintexts and $\epsilon$, the bias, be sufficiently small. Let $q^{(i)}$ be the probability that the following equation holds for a subkey candidate $K_n^{(i)}$ and a random variable X:
\[ F_n(X,K_n)[l_1, l_2,\dots,l_d] = F_n(X, K_n^{(i)})[l_1, l_2, \dots, l_d].\]
Then if $q^{(i)}$'s are independent, the success rate of the attack is
\[ \int_{x=-2\sqrt{N}\epsilon}^\infty \left( \prod_{K_n^{(i)} \neq K_n} \int_{-x-4\sqrt{N}\epsilon q^{(i)}}^{x+4\sqrt{N}\epsilon (1 - q^{(i)})} \frac{1}{\sqrt{2\pi}} e^{-y^2/2} dy \right) \frac{1}{\sqrt{2\pi}} e^{-x^2/2} dx \]
Where the product is taken over all subkey candidates except $K_n$.\\
\vspace{3mm}
Let $d = 1$ and $l_1 = 17$ in the above expression - this corresponds to our attack (distribution over guessing for one bit, $CR_{17}$ in the equation). The following table summarizes the sucess rates:
\begin{center}
\begin{tabular}{|c|r|r|r|r|}
\hline
N & 2$\epsilon^{-2}$ & $4\epsilon^{-2}$ & $8\epsilon^{-2}$ & $16\epsilon^{-2}$ \\
\hline
Success Rate & 0.486 & 0.785 & 0.967 & 0.999 \\
\hline
\end{tabular} 
\end{center}
\end{frame}

\begin{frame}
How many pairs do we need? Continued\\
\vspace{3mm}
Using the calculation from the previous slide, we see that for a 4-round DES, our equation holds with bias of about 1 in 4. Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of sucess.\\
\vspace{3mm}
For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
\end{frame}
